---
title: 'Assignment 10: Data Scraping'
author: "Iddrisu Sharu Deen"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, warning=FALSE, message=FALSE}
#1 
library(tidyverse); library(rvest); library(here); library(lubridate); library(purrr)

here()

mytheme <- theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(face = "italic", size = 12),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 8.5),
    legend.position = "bottom",
    legend.title = element_blank())

theme_set(mytheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website, warning=FALSE, message=FALSE}
#2 
webpage <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022")
webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data, warning=FALSE, message=FALSE}
#3 
water_system <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_system

PWSID <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

ownership <- webpage %>% 
 html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max_day_use_per_month <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max_day_use_per_month

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data, message=FALSE, warning=FALSE}
#4 
durham22.df <- data.frame("Month" = as.factor(c(1,5,9,2,6,10,3,7,11,4,8,12)),
                          "Year" = rep(2022, 12),
                          "Water System" = water_system,
                          "Ownership" = ownership,
                          "Max Daily Usage Per Month" = as.numeric(max_day_use_per_month)
                          )

durham22.df <- arrange(durham22.df, by_group = Month)

#5 
max_daily_withdrawals.plt <- ggplot(durham22.df, aes(x=Month, y=Max.Daily.Usage.Per.Month), group = 1) +
  geom_line(group=1) +
  geom_point()+
  labs(title = "Maximum Daily Withdrawals By Month",
       subtitle = "Year: 2022",
       caption = "Source: www.ncwater.org",
       x = "Month",
       y = "Max Daily WIthdrawals")
max_daily_withdrawals.plt

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function, warning=FALSE, message=FALSE}
#6. 
scrape.function <- function(year, pwsid_tag){
  # Get site content
  site.url <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                               pwsid_tag, '&year=', year))
  
  # Set variables
  water_system.tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
  pwsid.scrape.tag <- "td tr:nth-child(1) td:nth-child(5)"
  ownership.tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
  usage.tag <- "th~ td+ td"
  
  # Scrape data
  water_system.scrape <- site.url %>% html_nodes(water_system.tag) %>% html_text()
  pwsid.scrape <- site.url %>% html_nodes(pwsid.scrape.tag) %>% html_text()
  ownership.scrape <- site.url %>% html_nodes(ownership.tag) %>% html_text()
  usage.scrape <- site.url %>% html_nodes(usage.tag) %>% html_text()
  
  # Convert to dataframe
  withdrawals.df <- data.frame("Month" = factor(1:12),
                               "Year" = rep(year, 12),
                               "Max.Daily.Usage.Per.Month" = as.numeric(usage.scrape)) %>%
    mutate(Water.System = water_system.scrape,
           PWSID = pwsid.scrape,
           Ownership = ownership.scrape,
           Date = make_date(Year, Month, day = 1))
  
  # Pause
  Sys.sleep(1)
  
  # Output
  return(withdrawals.df)
}
```




7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, warning=FALSE, message=FALSE}
#7 
durham15.df <- scrape.function(2015, "03-32-010")

durham15.plt <- ggplot(durham15.df, aes(x = Month, y = Max.Daily.Usage.Per.Month)) +
  geom_point() +
  geom_line(group = 1) +
  labs(
    title = "Maximum Daily Withdrawals in Durham by Month",
    subtitle = "Year: 2015",
    caption = "Source: www.ncwater.org",
    x = "Month",
    y = "Max Daily Withdrawals"
  )
durham15.plt
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, warning=FALSE, message=FALSE}
#8 
asheville15.df <- scrape.function(2015, "01-11-010")

combined_data <- bind_rows(durham15.df, asheville15.df)

combined_plt <- ggplot(combined_data, aes(x = Month, y = Max.Daily.Usage.Per.Month, color = Water.System)) +
  geom_point() +
  geom_line(aes(group = Water.System)) +
  labs(
    title = "Comparison of Water Withdrawals in Asheville and Durham by Month",
    subtitle = "Year: 2015",
    caption = "Source: www.ncwater.org",
    x = "Month",
    y = "Max Daily Withdrawals"
  )
combined_plt
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data, warning=FALSE, message=FALSE }
#9 
years10_21 <- 2010:2021
asheville_pwsid = "01-11-010"

asheville10_21.df <- asheville10_21.df <- map2(years10_21, asheville_pwsid , scrape.function) %>%
  bind_rows()

asheville10_21.plt <- ggplot(asheville10_21.df, aes(x = Date, y = Max.Daily.Usage.Per.Month)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = 'loess', se = TRUE, color = "blue") +
  labs(
    title = "Maximum Daily Withdrawals in Asheville by Month",
    subtitle = "Years: 2010-2021",
    caption = "Source: www.ncwater.org",
    x = "Date",
    y = "Max Daily Withdrawals"
  )
asheville10_21.plt
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: There appears to be an overall slight positive trend in water usage in Asheville
