---
title: 'Assignment 2: Coding Basics'
author: "Iddrisu Sharu Deen"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

```{r number_sequence}
num_seq <- seq(from = 1, to = 30)
num_seq
```

2.  Compute the mean and median of this sequence.

```{r mean_&_median}
num_seq.mean <- mean(num_seq)
num_seq.mean

num_seq.median <- median(num_seq)
num_seq.median
```

3.  Ask R to determine whether the mean is greater than the median.

```{r compare_mean_median}
num_seq.mean > num_seq.median

```

4.  Insert comments in your code to describe what you are doing. 

```{r basics 1}
#1. I used the seq() function to generate the sequence of numbers with the from, to arguments denoting the start and end numbers, and saved the generated sequence in the variable num_seq.

#2. I passed num_seq to the mean() funtion to generate the mean, and saved the result in the variable num_seq.mean

#3. I passed num_seq to the median() funtion to generate the mean, and saved the result in the variable num_seq.median

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

```{r}
student_names <- c("John", "Jane", "Jill", "Jim")
test_scores <- c(70, 49, 65, 45)
passed <- ifelse(test_scores > 50, TRUE, FALSE)

```

6.  Label each vector with a comment on what type of vector it is.

```{r}
# student_names is a character vector
class(student_names)


# test_scores is a numeric vector
class(test_scores)


#passed is a logical vector
class(passed)
```

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

```{r}
student_performance <- data.frame(student_names, test_scores, passed)
student_performance
```

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
names(student_performance) <- c("Student Names", "Test Scores", "Passed")
student_performance
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frames can contain multpile datatypes whereas matrices can only contain one datatype. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

```{r}
myfunc <- function(x){
  ifelse(x > 50, TRUE, FALSE)}


```

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
myfunc(test_scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The -ifelse- option is better because it uses fewer lines of code compared to the -if- and -else-
